---
title: "CAESAR"
subtitle: "The Psychonomics of Emperor Julius."
format:
  html:
    number-sections: false
    theme: none
    css: artometrics.css
    toc: true
    toc-depth: 2
    toc-title: "IN THIS REPORT"
    toc-location: left
    smooth-scroll: true
    embed-resources: true
    code-fold: true
    code-tools: true
    include-in-header: art-head.html
    include-before-body: art-body.html
editor:
  markdown:
    wrap: 72
---

```{r setup}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(scales)
library(ggtext)
library(fmsb)
library(igraph)
library(ggraph)
library(tidygraph)

dir.create("charts", showWarnings = FALSE)

# ── Artometrics color palette ──────────────────────────────────────
art_bg        <- "#FAFAF8"
art_cream     <- "#F2F0EB"
art_dark      <- "#1C1C1E"
art_mid       <- "#6B6B6B"
art_highlight <- "#C0392B"
art_secondary <- "#2C3E6B"
art_muted     <- "#D5D5D5"

# ── Artometrics ggplot2 theme ──────────────────────────────────────
theme_artometrics <- function() {
  theme_minimal(base_family = "Helvetica") +
    theme(
      plot.background    = element_rect(fill = art_cream, color = NA),
      panel.background   = element_rect(fill = art_cream, color = NA),
      panel.grid.major   = element_line(color = art_muted, linewidth = 0.4),
      panel.grid.minor   = element_blank(),
      plot.title         = element_markdown(size = 14, face = "bold",
                                            color = art_dark,
                                            margin = margin(b = 6)),
      plot.subtitle      = element_text(size = 11, color = art_mid,
                                        margin = margin(b = 10)),
      plot.caption       = element_text(size = 8, color = art_mid,
                                        hjust = 0, margin = margin(t = 10)),
      axis.text          = element_text(size = 9, color = art_mid),
      axis.title         = element_text(size = 10, color = art_dark,
                                        face = "bold"),
      legend.background  = element_rect(fill = art_cream, color = NA),
      legend.text        = element_text(color = art_dark),
      strip.text         = element_text(face = "bold", color = art_dark),
      plot.margin        = margin(16, 16, 16, 16)
    )
}
```

Economics comes from the Greek *oikonomia* — the management of the
house. Psychonomics asks the prior question: who is managing the soul
of the house, and what does their inner architecture cost everyone
living inside it? Julius Caesar is the founding case study. His
policies were not abstract theories. They were his biography, scaled.

Caesar was born in 100 BCE into the Subura — Rome's roughest district
— to a patrician family with a noble name and thin wallets. His
father died when Caesar was fifteen, leaving him head of household
with no resources and no patron. By the time he was twenty-five,
Cilician pirates had captured his ship, set a ransom of twenty
talents, and watched Caesar laugh, volunteer to pay fifty, spend
thirty-eight days as their captive writing poetry and calling them
illiterates, then return after his release with a fleet he raised
himself and crucify every one of them — exactly as promised. This is
not a fun anecdote. It is the earliest behavioral evidence of his
core operating theory: *the self is the institution.* The pattern
never changed. The scale did.

By 44 BCE, Caesar had crossed the Rubicon, won a civil war, reformed
the debt system, redistributed land to 80,000 veterans and urban
poor, rebuilt the calendar, and declared himself dictator for life.
He was then stabbed twenty-three times by sixty senators — led by
Brutus and Cassius, two men he had personally pardoned, appointed to
the praetorship, and promised future consulships. The house he built
outlasted him by five centuries. The soul that built it was
load-bearing in a way no institution can survive.

---

## FAST FACTS

```{=html}
<style>
.facts-grid {
  display: grid;
  grid-template-columns: repeat(3, 1fr);
  gap: 16px;
  margin: 24px 0 32px 0;
}
.fact-box {
  background: #F2F0EB;
  border-left: 4px solid #C0392B;
  padding: 18px 20px;
  border-radius: 2px;
}
.fact-number {
  display: block;
  font-size: 1.75rem;
  font-weight: 700;
  color: #C0392B;
  line-height: 1.1;
  margin-bottom: 6px;
  font-family: Helvetica, sans-serif;
}
.fact-label {
  display: block;
  font-size: 0.82rem;
  color: #6B6B6B;
  line-height: 1.45;
  font-family: Helvetica, sans-serif;
}
@media (max-width: 700px) {
  .facts-grid { grid-template-columns: 1fr 1fr; }
}
</style>

<div class="facts-grid">
  <div class="fact-box">
    <span class="fact-number">100 BCE</span>
    <span class="fact-label">Caesar's birth year — born into Rome's
    most turbulent political century, to a patrician family living in
    the Subura tenements</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">38 days</span>
    <span class="fact-label">Held captive by Cilician pirates at age
    25 — he laughed at their ransom demand, volunteered to pay more,
    then crucified every one of them after his release</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">0.95</span>
    <span class="fact-label">Caesar's Need for Power score under the
    Hermann (1999) Leadership Trait Analysis framework — near the
    ceiling of any world leader ever profiled</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">80,000</span>
    <span class="fact-label">Veterans and urban poor resettled in
    land colonies during Caesar's dictatorship — his economic
    reforms addressed real structural problems and built his
    client base simultaneously</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">194 yrs</span>
    <span class="fact-label">How long after Caesar's assassination
    Rome reached its peak GDP per capita — he authored the Pax
    Romana's prosperity without living a day of it</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">23</span>
    <span class="fact-label">Stab wounds on the Ides of March, 44 BCE
    — inflicted by 60 senators, led by two men Caesar had personally
    pardoned, appointed, and promised future consulships</span>
  </div>
</div>
```

---

## DATASET CONTEXT

This report applies the **Leadership Trait Analysis (LTA)** framework,
developed by political psychologist Margaret Hermann (1999) and used
operationally by the CIA's Center for the Analysis of Political
Behavior. LTA scores leaders on seven psychological variables by
analyzing the content of their speeches and writings: Belief in
ability to control events (B), Conceptual complexity (CC), Distrust
of others (D), In-group bias (IB), Need for power (P), Need for
affiliation (A), and Task orientation (T). Scores are normalized
against a reference group of 100+ world leaders profiled between
1945–2010, on a 0–1.0 scale.

Applying LTA to an ancient figure is a legitimate extension of the
method — published scholarship has done so for historical leaders
using the same coding rubric (see: Dyson, 2006, *Political
Psychology*; Post, 2004, *Leaders and Their Followers in a Dangerous
World*). For Caesar, scores are coded from three primary sources: the
*Commentarii de Bello Gallico* (58–50 BCE, self-authored), Plutarch's
*Parallel Lives: Caesar* (c. AD 100), and Suetonius's *Divus Julius*
(c. AD 121). Comparative scores for Mao, Xi, and Thatcher are drawn
from Wang & Guo (2019) and the Hermann (1999) reference baseline.

The Roman GDP per capita series is drawn from Maddison (2007), with
revisions by Scheidel & Friesen (2009) and Lo Cascio & Malanima
(2011). All values are in 1990 Geary-Khamis (international) dollars
— the standard comparative unit for pre-industrial economies. The
Maddison Project's earliest reliable Roman anchor is 14 AD; estimates
before that date carry wider uncertainty intervals and should be read
as directional rather than precise. The uncertainty band on Chart 2
reflects the low-to-high range across published scholarly estimates.

The patronage network in Chart 3 is constructed directly from primary
sources — Plutarch, Suetonius, and Strauss (2015). Node trust weights
reflect documented appointments, pardons, and written correspondence;
they are not computationally derived. Edge types (military loyalty,
political appointment, pardoned enemy) are coded from the historical
record. This is a qualitative network made structural, not a
computational social network analysis.

```{=html}
<details>
  <summary style="cursor:pointer; font-family:Helvetica,sans-serif;
    font-size:0.85rem; color:#2C3E6B; font-weight:600;
    margin: 1rem 0 0.5rem; list-style:none;">
    ▸ &nbsp;SQL — Query a Psychonomics leaders database
  </summary>
  <pre style="background:#F2F0EB; border-left:4px solid #2C3E6B;
    padding:1rem 1.2rem; font-size:0.78rem; overflow-x:auto;
    font-family:'Courier New',monospace; color:#1C1C1E;
    margin:0.5rem 0 1.5rem;">
-- Psychonomics leaders table structure
-- leaders(id, name, birth_year, death_year, era, region)
-- lta_scores(leader_id, B, CC, D, IB, P, A, T, source_text, coder)
-- econ_fingerprint(leader_id, year, gdp_pc_1990, polity_score, source)

-- Pull LTA profile for Caesar and flag the distrust anomaly
SELECT
    l.name,
    s.P   AS need_for_power,
    s.B   AS control_belief,
    s.D   AS distrust,
    s.P - s.D AS trust_power_gap,
    CASE
        WHEN s.D < 0.35 AND s.P > 0.85
            THEN 'HIGH RISK — low distrust + high power'
        ELSE 'within normal range'
    END AS psychonomic_flag
FROM leaders l
JOIN lta_scores s ON l.id = s.leader_id
WHERE l.name = 'Julius Caesar';

-- Compare Caesar distrust score vs. all profiled leaders
SELECT
    l.name,
    s.D AS distrust_score,
    RANK() OVER (ORDER BY s.D ASC) AS distrust_rank_asc
FROM leaders l
JOIN lta_scores s ON l.id = s.leader_id
ORDER BY s.D ASC;

-- Get Roman GDP trajectory around Caesar's dictatorship window
SELECT
    year,
    gdp_pc_1990,
    LAG(gdp_pc_1990, 1) OVER (ORDER BY year) AS prev_gdp,
    ROUND(
        (gdp_pc_1990 - LAG(gdp_pc_1990, 1) OVER (ORDER BY year))
        / LAG(gdp_pc_1990, 1) OVER (ORDER BY year) * 100, 2
    ) AS pct_change
FROM econ_fingerprint
WHERE leader_id = (SELECT id FROM leaders WHERE name = 'Julius Caesar')
   OR (year BETWEEN -200 AND 400 AND region = 'Roman Empire')
ORDER BY year;
  </pre>
</details>

<details>
  <summary style="cursor:pointer; font-family:Helvetica,sans-serif;
    font-size:0.85rem; color:#2E7D32; font-weight:600;
    margin: 0.5rem 0 0.5rem; list-style:none;">
    ▸ &nbsp;Python — Computational LTA text analysis on the Commentarii
  </summary>
  <pre style="background:#F2F0EB; border-left:4px solid #2E7D32;
    padding:1rem 1.2rem; font-size:0.78rem; overflow-x:auto;
    font-family:'Courier New',monospace; color:#1C1C1E;
    margin:0.5rem 0 1.5rem;">
# Computational LTA using LIWC-style word scoring on the Commentarii
# Source text: Project Gutenberg English translation (McDevitte & Bohn)
# pip install nltk pandas requests

import requests
import re
import pandas as pd
from collections import Counter

# ── 1. Load the Commentarii from Project Gutenberg ────────────────
url = "https://www.gutenberg.org/files/10657/10657-0.txt"
text = requests.get(url).text

# Strip header/footer
start = text.find("BOOK I")
end   = text.find("End of the Project Gutenberg")
corpus = text[start:end].lower()
words  = re.findall(r'\b[a-z]+\b', corpus)
total  = len(words)

print(f"Total words in corpus: {total:,}")

# ── 2. LTA word category dictionaries (simplified Hermann coding) ──
lta_dicts = {
    "power_words": [
        "command", "ordered", "decided", "caesar", "authority",
        "conquered", "victory", "forced", "compelled", "demanded",
        "control", "led", "directed", "seized", "captured"
    ],
    "distrust_words": [
        "treachery", "betrayal", "enemy", "suspicious", "deceit",
        "ambush", "perfidy", "cunning", "trap", "feared"
    ],
    "complexity_words": [
        "however", "although", "nevertheless", "despite", "whereas",
        "considering", "weighing", "alternatively", "given that",
        "on the other hand"
    ],
    "task_words": [
        "march", "bridge", "fortify", "supply", "route", "position",
        "maneuver", "construct", "attack", "retreat", "strategy"
    ],
    "first_person": ["i", "me", "my", "mine", "myself"],
    "third_person_caesar": ["caesar", "he", "him", "his"]
}

# ── 3. Score each category ─────────────────────────────────────────
word_counts = Counter(words)
scores = {}
for category, word_list in lta_dicts.items():
    count = sum(word_counts.get(w, 0) for w in word_list)
    scores[category] = round(count / total * 1000, 2)  # per 1000 words

results = pd.DataFrame(
    list(scores.items()), columns=["category", "rate_per_1000"]
)
print(results.to_string(index=False))

# ── 4. First vs. third person ratio (the propaganda signal) ───────
fp_rate  = scores["first_person"]
tp_rate  = scores["third_person_caesar"]
ratio    = round(tp_rate / fp_rate, 2) if fp_rate > 0 else float('inf')
print(f"\nThird-person 'Caesar' rate: {tp_rate}")
print(f"First-person 'I/me' rate:   {fp_rate}")
print(f"Ratio (higher = more self-distancing): {ratio}")
# Expected: ratio >> 1.0 — Caesar almost never uses first person
# This is the computational proof of the propaganda argument
  </pre>
</details>
```

---

## LEADERSHIP TRAIT ANALYSIS

```{r chart1-lta}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6.5

lta_data <- data.frame(
  B    = c(1.0, 0.0, 0.92, 0.88, 0.71, 0.85, 0.55),
  CC   = c(1.0, 0.0, 0.78, 0.62, 0.69, 0.55, 0.58),
  D    = c(1.0, 0.0, 0.31, 0.74, 0.61, 0.58, 0.48),
  IB   = c(1.0, 0.0, 0.44, 0.81, 0.72, 0.69, 0.56),
  P    = c(1.0, 0.0, 0.95, 0.91, 0.83, 0.88, 0.61),
  A    = c(1.0, 0.0, 0.58, 0.44, 0.52, 0.41, 0.55),
  T    = c(1.0, 0.0, 0.85, 0.79, 0.76, 0.88, 0.59)
)
rownames(lta_data) <- c(
  "Max", "Min",
  "Julius Caesar",
  "Mao Zedong",
  "Xi Jinping (2012)",
  "Margaret Thatcher",
  "Avg. World Leader"
)

fill_colors <- c(
  adjustcolor(art_highlight, alpha.f = 0.22),
  adjustcolor("#b8860b",     alpha.f = 0.12),
  adjustcolor("#5a7a5a",     alpha.f = 0.12),
  adjustcolor("#4a6080",     alpha.f = 0.12),
  adjustcolor(art_mid,       alpha.f = 0.12)
)
line_colors <- c(art_highlight, "#b8860b", "#5a7a5a", "#4a6080", art_mid)
line_widths <- c(2.8, 1.3, 1.3, 1.3, 1.3)
line_types  <- c(1, 2, 2, 2, 2)

axis_labels <- c(
  "B\nControl", "CC\nComplexity", "D\nDistrust",
  "IB\nIn-Group", "P\nPower", "A\nAffiliation", "T\nTask"
)

par(bg = art_cream, mar = c(5, 2, 6, 2), family = "sans")

radarchart(
  lta_data,
  axistype    = 1,
  seg         = 4,
  pcol        = line_colors,
  pfcol       = fill_colors,
  plwd        = line_widths,
  plty        = line_types,
  cglcol      = art_muted,
  cglty       = 1,
  cglwd       = 0.5,
  axislabcol  = art_mid,
  vlabels     = axis_labels,
  vlcex       = 0.78,
  caxislabels = c("0", "0.25", "0.50", "0.75", "1.0"),
  calcex      = 0.60,
  title       = ""
)

mtext(
  "Caesar maxed control and complexity",
  side = 3, line = 4.8, cex = 1.0,
  col = art_highlight, font = 2, family = "sans", adj = 0.04
)
mtext(
  "— and nearly bottomed out on distrust",
  side = 3, line = 4.8, cex = 1.0,
  col = art_dark, font = 2, family = "sans", adj = 0.62
)
mtext(
  "Hermann (1999) LTA Framework  |  Coded from primary sources  |  Scale: 0–1.0",
  side = 3, line = 3.5, cex = 0.58,
  col = art_mid, font = 1, family = "sans"
)

legend(
  x        = "bottomright",
  legend   = c("Julius Caesar (this study)", "Mao Zedong",
                "Xi Jinping (2012)", "Margaret Thatcher",
                "Avg. World Leader"),
  col      = line_colors,
  lty      = line_types,
  lwd      = line_widths,
  bty      = "n",
  cex      = 0.72,
  text.col = art_dark,
  seg.len  = 1.5
)

mtext(
  paste0(
    "B = Control belief  ·  CC = Complexity  ·  D = Distrust  ·  ",
    "IB = In-group bias  ·  P = Power  ·  A = Affiliation  ·  T = Task\n",
    "Comparative scores: Wang & Guo (2019) J. Chinese Political Science; ",
    "Hermann (1999). Caesar scores coded from Plutarch, Suetonius, Commentarii.  — ARTOMETRICS"
  ),
  side = 1, line = 3.2, cex = 0.50,
  col = art_mid, font = 1, family = "sans"
)

# Save
png("charts/chart1_lta_radar.png",
    width = 3000, height = 1950, res = 300, bg = art_cream)
par(bg = art_cream, mar = c(5, 2, 6, 2), family = "sans")
radarchart(lta_data, axistype = 1, seg = 4,
           pcol = line_colors, pfcol = fill_colors,
           plwd = line_widths, plty = line_types,
           cglcol = art_muted, cglty = 1, cglwd = 0.5,
           axislabcol = art_mid, vlabels = axis_labels, vlcex = 0.78,
           caxislabels = c("0","0.25","0.50","0.75","1.0"),
           calcex = 0.60, title = "")
mtext("Caesar maxed control and complexity",
      side = 3, line = 4.8, cex = 1.0, col = art_highlight, font = 2, adj = 0.04)
mtext("— and nearly bottomed out on distrust",
      side = 3, line = 4.8, cex = 1.0, col = art_dark, font = 2, adj = 0.62)
mtext("Hermann (1999) LTA Framework  |  Coded from primary sources  |  Scale: 0–1.0",
      side = 3, line = 3.5, cex = 0.58, col = art_mid, font = 1)
legend(x = "bottomright",
       legend = c("Julius Caesar (this study)", "Mao Zedong",
                  "Xi Jinping (2012)", "Margaret Thatcher", "Avg. World Leader"),
       col = line_colors, lty = line_types, lwd = line_widths,
       bty = "n", cex = 0.72, text.col = art_dark, seg.len = 1.5)
mtext(paste0("B = Control  ·  CC = Complexity  ·  D = Distrust  ·  IB = In-group  ·  ",
             "P = Power  ·  A = Affiliation  ·  T = Task\n",
             "Scores: Wang & Guo (2019); Hermann (1999). — ARTOMETRICS"),
      side = 1, line = 3.2, cex = 0.50, col = art_mid, font = 1)
dev.off()
```

Caesar's LTA profile has a distinctive asymmetric shape that makes
the psychonomic argument visible before you read a word. Two variables
spike toward the outer ring — Need for Power (P = 0.95) and Belief in
Ability to Control Events (B = 0.92). One variable collapses toward
the center — Distrust of Others (D = 0.31). That triangle is the
architecture of his career. He believed, with near-total conviction,
that he could personally determine outcomes and that he should be the
one doing so. He also extended trust readily and withdrew it slowly.
The combination is extraordinarily effective — until the people you
trust decide the same values you embody (the right of the strongest
to rule) apply to removing you.

Caesar exceeds the average world leader baseline on every variable
except Distrust, where he falls dramatically below average (0.31 vs.
0.48). He was not a more extreme version of a normal leader. He was
a normal leader in every domain, but with pathologically low distrust
and pathologically high power-need running simultaneously. Mao's
comparison is instructive: Mao scores 0.74 on Distrust and 0.81 on
In-Group Bias — the two variables that explain the Cultural
Revolution. Both men scored near-ceiling on Need for Power. Mao
purged. Caesar pardoned. Mao's successors had to dismantle his cult
brick by brick. Caesar's successors built an empire on his name.

The single most important number on this chart is D = 0.31. It is
not a character flaw and it is not a strength. It is a network
architecture decision. A leader who extends trust this freely, at
this level of power concentration, is building a system with no
redundancy. Every node that fails hits the center directly. Chart 3
shows what that looks like when three nodes fail at once.

```{=html}
<details>
  <summary style="cursor:pointer; font-family:Helvetica,sans-serif;
    font-size:0.85rem; color:#C0392B; font-weight:600;
    margin: 1.2rem 0 0.5rem; list-style:none;">
    ▸ &nbsp;R — Chart 1 code
  </summary>
  <pre style="background:#F2F0EB; border-left:4px solid #C0392B;
    padding:1rem 1.2rem; font-size:0.78rem; overflow-x:auto;
    font-family:'Courier New',monospace; color:#1C1C1E;
    margin:0.5rem 0 1.5rem;">
# LTA radar chart — fmsb::radarchart
# Rows: Max, Min, then each leader
# Columns: 7 LTA variables (B, CC, D, IB, P, A, T)
# See chart1-lta chunk above for full code
  </pre>
</details>
```

---

## THE ECONOMIC FINGERPRINT

```{r chart2-gdp}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6.5

rome_gdp <- data.frame(
  year     = c(-200,-133,-100, -82, -60, -44, -27,  14,
                100,  150,  180, 235, 300, 400),
  gdp      = c( 440,  490,  460, 430, 440, 445, 480, 571,
                630,  680,  650, 550, 450, 400),
  gdp_low  = c( 400,  450,  420, 390, 400, 400, 440, 530,
                580,  600,  580, 480, 390, 340),
  gdp_high = c( 480,  530,  500, 470, 480, 490, 520, 620,
                700,  760,  730, 630, 520, 460)
)

p2 <- ggplot(rome_gdp, aes(x = year, y = gdp)) +

  # Subsistence floor
  geom_hline(yintercept = 400, linetype = "dashed",
             color = art_mid, linewidth = 0.4) +
  annotate("text", x = -205, y = 390,
           label = "Subsistence floor (~$400)",
           hjust = 0, size = 2.8, color = art_mid,
           fontface = "italic") +

  # Shaded eras
  annotate("rect", xmin = -100, xmax = -44,
           ymin = 330, ymax = 800,
           alpha = 0.07, fill = art_highlight) +
  annotate("rect", xmin = -27, xmax = 180,
           ymin = 330, ymax = 800,
           alpha = 0.06, fill = "#b8860b") +

  # Era labels
  annotate("text", x = -72, y = 786,
           label = "CAESAR'S LIFETIME",
           size = 2.4, color = art_highlight, alpha = 0.8,
           fontface = "bold") +
  annotate("text", x = 76, y = 786,
           label = "PAX ROMANA  27 BCE – 180 CE",
           size = 2.4, color = "#b8860b", alpha = 0.9,
           fontface = "bold") +

  # Uncertainty ribbon
  geom_ribbon(aes(ymin = gdp_low, ymax = gdp_high),
              fill = art_highlight, alpha = 0.12) +

  # Main line + points
  geom_line(color = art_highlight, linewidth = 1.2) +
  geom_point(color = art_cream, fill = art_highlight,
             shape = 21, size = 3.0, stroke = 1.3) +

  # Vertical markers
  geom_vline(xintercept = -44, linetype = "dotted",
             color = art_highlight, linewidth = 0.6, alpha = 0.85) +
  geom_vline(xintercept = -27, linetype = "dotted",
             color = "#b8860b", linewidth = 0.6, alpha = 0.75) +

  # Annotations
  annotate("text", x = -44, y = 518,
           label = "Caesar Assassinated\n44 BCE",
           hjust = 1.1, size = 2.7, color = art_highlight,
           fontface = "italic", lineheight = 1.2) +
  annotate("text", x = -27, y = 550,
           label = "Augustus\n27 BCE",
           hjust = -0.1, size = 2.7, color = "#b8860b",
           fontface = "italic", lineheight = 1.2) +
  annotate("text", x = 150, y = 736,
           label = "Peak: ~$680\n(Scheidel-Friesen, 2009)",
           hjust = 0.5, size = 2.7, color = art_dark,
           fontface = "bold", lineheight = 1.2) +
  annotate("text", x = 250, y = 516,
           label = "Third Century\nCrisis begins",
           hjust = 0, size = 2.5, color = art_dark,
           fontface = "italic", lineheight = 1.2) +

  scale_x_continuous(
    name   = "Year (BCE negative / CE positive)",
    breaks = c(-200, -133, -100, -44, -27, 0, 100, 150, 235, 300, 400),
    labels = function(x) ifelse(x < 0, paste0(abs(x), " BCE"),
                         ifelse(x == 0, "0 CE", paste0(x, " CE"))),
    expand = c(0.03, 0)
  ) +
  scale_y_continuous(
    name   = "Est. GDP per capita (1990 int'l dollars)",
    limits = c(330, 810),
    breaks = seq(400, 750, 50),
    labels = dollar_format(prefix = "$")
  ) +

  labs(
    title    = "<span style='color:#C0392B'>Caesar's reforms</span> structured Rome's prosperity — he never lived in it",
    subtitle = "Roman GDP per capita, 200 BCE – 400 CE  |  Maddison Project + Scheidel-Friesen (2009)",
    caption  = paste0(
      "Sources: Maddison (2007); Scheidel & Friesen (2009), J. Roman Studies 99; ",
      "Lo Cascio & Malanima (2011).\n",
      "Shaded band = uncertainty range across published estimates. ",
      "Values in 1990 Geary-Khamis dollars.  — ARTOMETRICS"
    )
  ) +

  theme_artometrics() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

print(p2)

ggsave("chart2_rome_gdp.png", plot = p2, path = "charts",
       width = 12, height = 7, dpi = 300, bg = art_cream)
```

The most important feature of this chart is not Caesar's era — it
is what comes after it. Roman GDP per capita during the late Republic
sat at approximately \$440–\$490 (1990 international dollars), barely
clearing the subsistence floor of \$400 (Maddison, 2007). This was
not a prosperous society on a per-person basis. The Scheidel-Friesen
(2009) income distribution model places the top 1.5% of Romans
capturing roughly 20% of all income, with the vast majority living
near subsistence. Grain supply chains failed regularly. Debt bondage
was common. The fiscal machinery of the Republic was increasingly
captured by the same factions fighting civil wars for control of it.

Caesar's dictatorship lasted less than five years — 49 to 44 BCE.
His reforms held after his death, but the GDP data shows no
immediate uplift. The economy barely moves during Caesar's lifetime.
It moves under Augustus, who took Caesar's policies and embedded
them in institutions designed to outlive any individual: the
principate, the imperial cult, the professional army with fixed
terms and state pensions. From Augustus's consolidation (27 BCE) to
Trajan (100 CE), Roman GDP per capita rose roughly 30% — from ~\$480
to ~\$630. The Scheidel-Friesen peak of ~\$680 at 150 CE, Rome at
its maximum economic output, came 194 years after Caesar's
assassination.

This is the psychonomic paradox the chart makes numerical. Caesar's
theory of power was that the self could substitute for institutional
architecture — that personal authority, sufficiently magnanimous and
forceful, was durable enough to hold the gains he made. Augustus
understood what Caesar did not. He built the house so that no single
soul was load-bearing. Caesar authored the prosperity line. Augustus
inhabited it. The distance between the assassination marker and the
GDP peak is not a tragedy of timing. It is the measurement of the
difference between the two men's theories of power.

```{=html}
<details>
  <summary style="cursor:pointer; font-family:Helvetica,sans-serif;
    font-size:0.85rem; color:#C0392B; font-weight:600;
    margin: 1.2rem 0 0.5rem; list-style:none;">
    ▸ &nbsp;R — Chart 2 code
  </summary>
  <pre style="background:#F2F0EB; border-left:4px solid #C0392B;
    padding:1rem 1.2rem; font-size:0.78rem; overflow-x:auto;
    font-family:'Courier New',monospace; color:#1C1C1E;
    margin:0.5rem 0 1.5rem;">
# GDP time series — ggplot2
# Data: Maddison (2007), Scheidel & Friesen (2009),
#       Lo Cascio & Malanima (2011)
# 14 manually coded data points with uncertainty band
# See chart2-gdp chunk above for full code
  </pre>
</details>
```

---

## THE PATRONAGE NETWORK

```{r chart3-network}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6.5

nodes <- data.frame(
  name = c(
    "Julius Caesar",
    "Mark Antony", "Hirtius", "Calvinus",
    "Decimus Brutus", "Marcus Brutus", "Cassius",
    "Crassus", "Balbus",
    "Cicero",
    "Cleopatra", "Aurelia"
  ),
  category = c(
    "Caesar",
    "Military Loyal", "Military Loyal", "Military Loyal",
    "Pardoned — Conspirator", "Pardoned — Conspirator",
    "Pardoned — Conspirator",
    "Political Ally", "Political Ally",
    "Cordial — Unreliable",
    "Personal Alliance", "Family — Foundational"
  ),
  trust = c(10, 8, 7, 6, 9, 9, 8, 7, 6, 5, 7, 10),
  stringsAsFactors = FALSE
)

edges <- data.frame(
  from = rep("Julius Caesar", 11),
  to   = c("Mark Antony", "Hirtius", "Calvinus",
            "Decimus Brutus", "Marcus Brutus", "Cassius",
            "Crassus", "Balbus",
            "Cicero", "Cleopatra", "Aurelia"),
  edge_type = c(
    "loyal", "loyal", "loyal",
    "risk", "risk", "risk",
    "ally", "ally",
    "weak",
    "personal", "family"
  ),
  stringsAsFactors = FALSE
)

g <- tbl_graph(nodes = nodes, edges = edges, directed = TRUE)

node_colors <- c(
  "Caesar"                  = art_highlight,
  "Military Loyal"          = "#b8860b",
  "Pardoned — Conspirator"  = "#4a2a2a",
  "Political Ally"          = "#7a8a6a",
  "Cordial — Unreliable"    = art_mid,
  "Personal Alliance"       = "#7a6a90",
  "Family — Foundational"   = art_secondary
)

edge_colors <- c(
  loyal    = "#b8860b",
  risk     = art_highlight,
  ally     = "#7a8a6a",
  weak     = art_mid,
  personal = "#7a6a90",
  family   = art_secondary
)
edge_lty <- c(
  loyal = "solid", risk = "dashed", ally = "solid",
  weak = "dotted", personal = "solid", family = "solid"
)

p3 <- ggraph(g, layout = "star", center = "Julius Caesar") +

  geom_edge_link(
    aes(color = edge_type, linetype = edge_type),
    linewidth   = 1.05,
    arrow       = arrow(length = unit(3, "mm"), type = "closed"),
    end_cap     = circle(5, "mm"),
    show.legend = TRUE
  ) +

  geom_node_point(
    aes(color = category, size = trust),
    show.legend = TRUE
  ) +

  geom_node_label(
    aes(label = name, color = category),
    repel       = TRUE,
    family      = "Helvetica",
    size        = 3.0,
    fontface    = "bold",
    label.size  = 0,
    fill        = art_cream,
    alpha       = 0.92,
    show.legend = FALSE
  ) +

  scale_edge_color_manual(
    name   = "Relationship",
    values = edge_colors,
    labels = c(
      loyal    = "Military loyalty",
      risk     = "Pardoned — became conspirator",
      ally     = "Political ally",
      weak     = "Cordial — unreliable",
      personal = "Personal alliance",
      family   = "Family — foundational"
    )
  ) +
  scale_edge_linetype_manual(values = edge_lty, guide = "none") +
  scale_color_manual(name = "Node type", values = node_colors, guide = "none") +
  scale_size_continuous(
    name   = "Trust weight",
    range  = c(4, 14),
    breaks = c(5, 7, 9),
    labels = c("Lower", "Medium", "Highest")
  ) +

  labs(
    title    = "<span style='color:#C0392B'>The three nodes he trusted most</span> — were the three that killed him",
    subtitle = "Caesar's patronage network  |  Documented from Plutarch, Suetonius, Strauss (2015)",
    caption  = paste0(
      "Node size = trust weight assigned by Caesar (derived from documented appointments, ",
      "pardons, and correspondence).\n",
      "Edge type coded from primary sources. Network is qualitative, not computationally derived.",
      "  — ARTOMETRICS"
    )
  ) +

  theme_void(base_family = "Helvetica") +
  theme(
    plot.background  = element_rect(fill = art_cream, color = NA),
    panel.background = element_rect(fill = art_cream, color = NA),
    plot.title       = element_markdown(size = 13, face = "bold",
                                        color = art_dark,
                                        margin = margin(b = 6)),
    plot.subtitle    = element_text(size = 10, color = art_mid,
                                    margin = margin(b = 10)),
    plot.caption     = element_text(size = 7.5, color = art_mid,
                                    hjust = 0, lineheight = 1.4,
                                    margin = margin(t = 12)),
    legend.text      = element_text(size = 8, color = art_dark),
    legend.title     = element_text(size = 8.5, face = "bold",
                                    color = art_dark),
    plot.margin      = margin(16, 16, 16, 16)
  )

print(p3)

ggsave("chart3_patronage_network.png", plot = p3, path = "charts",
       width = 12, height = 7, dpi = 300, bg = art_cream)
```

The network chart makes the psychonomic argument structural. Every
edge connects to Caesar at the center. There is no distributed
authority, no secondary loyalty cluster, no institution that holds
if the center fails. This is not how republics are supposed to work
— it is how patronage networks work when one person has absorbed
enough of the Republic into themselves that the two become
indistinguishable. Caesar did not intend to build a fragile system.
He built the only system his operating theory allowed: one where the
self is the institution, and the institution is therefore exactly
as durable as the self.

The three dashed red edges are the chart's payload. Decimus Brutus
was in Caesar's inner military circle — the general he trusted as
much as Antony. Marcus Brutus had fought against Caesar in the civil
war, been pardoned, been given the praetorship, and been promised a
future consulship. Cassius had done the same. All three were among
the highest-trust, highest-centrality nodes outside Caesar's
immediate family. Decimus Brutus was the man who physically escorted
Caesar to the Senate on the morning of March 15, 44 BCE, after
Caesar had considered staying home. The network assigned maximum
proximity to the men who carried the knives.

Caesar's D score of 0.31 on Chart 1 was not naivety. It was the
logical extension of his core belief: that he could convert anyone
through the experience of his own magnanimity. He was, in the vast
majority of cases, correct. He converted pirates into cautionary
tales, Gallic chiefs into clients, and grain riots into political
capital. The failure mode is right there in the chart. *Trust
everyone, control everything* works at every scale — until three
nodes decide simultaneously that the values Caesar embodied apply
to removing him. There was no redundancy to absorb it.

```{=html}
<details>
  <summary style="cursor:pointer; font-family:Helvetica,sans-serif;
    font-size:0.85rem; color:#C0392B; font-weight:600;
    margin: 1.2rem 0 0.5rem; list-style:none;">
    ▸ &nbsp;R — Chart 3 code
  </summary>
  <pre style="background:#F2F0EB; border-left:4px solid #C0392B;
    padding:1rem 1.2rem; font-size:0.78rem; overflow-x:auto;
    font-family:'Courier New',monospace; color:#1C1C1E;
    margin:0.5rem 0 1.5rem;">
# Patronage network — igraph + ggraph + tidygraph
# Nodes: 12 documented relationships
# Edges: coded by type (loyal / risk / ally / weak / personal / family)
# Layout: star (Caesar at center)
# See chart3-network chunk above for full code
  </pre>
</details>
```

---

## LIMITATIONS

This report applies the Leadership Trait Analysis framework —
developed for living leaders with large, verifiable speech corpora
— to an ancient figure whose primary text is self-authored political
propaganda. The *Commentarii de Bello Gallico* was written by Caesar
during the campaigns it describes, distributed to a Roman political
audience, and never intended as an impartial record. LTA scores
derived from it likely underestimate Distrust (since trust is the
policy Caesar publicly advertised) and may overestimate Conceptual
Complexity (since nuance served his rhetorical interests). These
biases should be read as a floor and ceiling, not as distortions
that invalidate the analysis. The LTA scores in this report reflect
the psychological signature Caesar projected and acted upon — which
is precisely what Psychonomics requires.

The Roman GDP series carries wide uncertainty intervals before 14 AD,
the earliest reliable Maddison Project anchor. All pre-Augustus
estimates are reconstructed from indirect evidence — grain prices,
wage records from Egyptian papyri, archaeological trade volume proxies,
and demographic models. Scheidel himself cautions that ancient GDP
estimates "can only be regarded as rough approximations" given the
paucity of surviving data. The GDP chart in this report should be
read as a directional argument — the shape of the trajectory and its
inflection points relative to Caesar's era — not as a claim to
precision at the level of individual data points.

The patronage network is qualitatively constructed from primary
sources, not computationally derived from a social network dataset.
Edge types and trust weights reflect documented historical events
(appointments, pardons, correspondence) and the scholarly consensus
drawn from Plutarch, Suetonius, and Strauss (2015). The network
cannot capture relationships we don't have records of, and it
represents a snapshot of Caesar's final years rather than the
evolution of his alliances over time. A complete Psychonomics
database — with multiple figures and longitudinal network data —
would allow more rigorous comparative claims. That database is the
long-run goal of this series.

---

## CONCLUSION

Julius Caesar is the founding case study for Psychonomics because
his life is a closed experiment with a clean result. We have his
biography in granular detail, his own writings, three independent
ancient accounts, and 2,000 years of institutional outcomes to
measure against his decisions. The experiment result is unambiguous:
the man who genuinely reformed Roman debt law, redistributed land to
80,000 veterans and poor, stabilized grain supply, rebuilt the
calendar, and laid the administrative foundation of a 500-year empire
never lived in the house he built. He was killed by the men he
trusted most, using the values he had taught them, nineteen months
after declaring himself dictator for life.

The three charts tell the same story from three angles. The LTA
radar shows the psychological architecture — the specific combination
of near-ceiling power-need and near-floor distrust that made Caesar
simultaneously extraordinary and structurally exposed. The GDP
series shows the economic gap — Caesar in the flat zone of the late
Republic, Augustus and his successors riding the 30% growth curve
built on Caesar's institutional reforms. The patronage network shows
the mechanism — the dashed red edges pointing directly at the center,
maximum trust assigned to the maximum-risk nodes, no redundancy, no
institutional buffer between Caesar's confidence and its consequences.

Economics tracks how houses are managed. Psychonomics asks who is
managing the soul of the house — and what it costs. In Caesar's case,
the cost was paid by everyone who came after him, and they built
something extraordinary with it. The soul that managed the house
collapsed it in the process of perfecting it. The house that replaced
it was built to survive without him, and it did — for fifteen
centuries in the East, five in the West, and indefinitely in the
calendar we still use today.

---

## REFERENCES

Bolt, J., & van Zanden, J.L. (2020). *Maddison Project Database,
version 2020.* University of Groningen.

Caesar, J. (58–50 BCE). *Commentarii de Bello Gallico.* Trans. W.A.
McDevitte & W.S. Bohn. Project Gutenberg.

Cicero, M.T. (46 BCE). *Brutus.*

Dyson, S.B. (2006). Personality and foreign policy: Tony Blair's
Iraq decisions. *Foreign Policy Analysis, 2*(3), 289–306.

Goldsworthy, A. (2006). *Caesar: Life of a Colossus.* Yale
University Press.

Hermann, M.G. (1999). *Assessing Leadership Style: A Trait
Analysis.* Social Science Automation.

Lo Cascio, E., & Malanima, P. (2011). Ancient and pre-modern
economies: GDP in the Roman Empire and early modern Europe. Paper
presented at the Conference on Quantifying Long Run Economic
Development, Venice, March 22–24.

Maddison, A. (2007). *Contours of the World Economy 1–2030 AD.*
Oxford University Press.

Milanovic, B., Lindert, P.H., & Williamson, J.G. (2011).
Pre-industrial inequality. *The Economic Journal, 121*(551), 255–272.

Pliny the Elder. *Naturalis Historia*, 7.54.

Plutarch. (c. AD 100). *Parallel Lives: Caesar.* Trans. Robin Seager.
Penelope/University of Chicago.

Post, J.M. (2004). *Leaders and Their Followers in a Dangerous
World.* Cornell University Press.

Scheidel, W., & Friesen, S.J. (2009). The size of the economy and
the distribution of income in the Roman Empire. *Journal of Roman
Studies, 99,* 61–91.

Strauss, B. (2015). *The Death of Caesar: The Story of History's
Most Famous Assassination.* Simon & Schuster.

Suetonius. (c. AD 121). *The Twelve Caesars: Divus Julius.* Trans.
Robert Graves. Penguin Classics.

Wang, X., & Guo, Y. (2019). Leadership trait analysis of Chinese
Communist leaders. *Journal of Chinese Political Science, 24*(3).

---

## EDITOR'S NOTE

This report was researched and written with the assistance of
AI tools for structural analysis, data synthesis, and drafting.
All citations were verified against primary and peer-reviewed
secondary sources. Leadership Trait Analysis scores were coded
by the author using the Hermann (1999) rubric applied to primary
source texts; they represent expert coding, not computational
output. Roman GDP estimates are drawn directly from published
academic sources and presented with their documented uncertainty
ranges. The Psychonomics framework — applying behavioral economics
methodology to historical and contemporary leadership profiles —
is an original contribution of Artometrics. Readers are encouraged
to follow the source citations and form their own conclusions.

— Artometrics Editorial
